### 2013-04-19 article.tex
Further proof read manuscript
Signed-off-by: Michael Barton <mail@michaelbarton.me.uk>

Manually editing a genome sequence or annotation is a non-trivial problem
requiring effort on a researcher's part. The large size of FASTA sequence files
or general feature format (GFF) annotation files makes changing these in a text
editor difficult. Tasks such as organising project files, adding additional
sequences to an assembly, updating incorrect annotations or preparing files for
database deposition all divert effort from the main goal of analysing a genome.
As a result genome projects have a steep learning curve for researchers
embarking on genome sequencing for the first time, while still requiring effort
for experienced researchers.

@@

During the process of sequencing several \emph{Pseudomonas fluorescens} strains
we developed a tool, called ``Genomer'', to ease and simplify many of the
common steps in our genome project. Several graphical user interface tools are
available to work with genome projects \cite{tanaka2006, wilkinson2002,
lopez2011, carver2012, gordon2003}. Genomer instead provides a simple, command
line interface for genome finishing allowing scriptable automation and
reproducible results. We have open-sourced this tool under the MIT license and
here we describe its application in a microbial genome project.

@@

Genomer uses these two libraries to create a plugin system where third-party
software can be written and included. Plugin creators need only to prefix the
name of their plugin with \verb+genomer-plugin-+ and upload this to
rubygems.org or make the plugin available as a public git repository. A plugin
is specified within a Genomer project in a file named `Gemfile.' The
\verb+bundle update+ command will then automatically download and install the
required plugins.

Genomer is implemented as a command line tool and tested on both Mac OS X and
Linux systems. The latest version of genomer can installed from the Rubygems
package management system with the command \verb+gem install genomer+. Genomer
is based on our previously described ``scaffolder'' software and is invoked on
the command line with the command \verb+genomer+. Genomer provides extensive
documentation via UNIX manual pages and on the genomer website at
http://next.gs.

@@

  \item[Mapping of annotations onto the assembled sequence.]{Genomer maps the
  coordinates of contig annotations to respective positions in the draft genome
  sequence. This allows the scaffold file to be continuously improved and
  updated even after annotation. For instance, additional PCR sequences can be
  used to close gaps in the assembly without reannotation or manually updating
  annotation coordinates.}

  \item[Generation of files for submission to GenBank.]{Submitting a genome
  sequence to GenBank database requires generating specific files. These may
  include a FASTA file of the draft sequence, a table of annotations, a FASTA
  file containing the individual contig sequences, and `a golden path' (AGP)
  \cite{agp-spec} file describing the placement of contigs. Genomer automates
  the generation of all these files from the scaffold file and corresponding
  GFF3 file.}

  \item[A stable interface and streamlined install process]{Genomer provides a
  plugin system built using the existing Ruby package management system:
  RubyGems and Bundler. This allows genomer and its plugins to be automatically
  installed without requiring manual downloading and compilation by the user.
  This eliminates a common problem in bioinformatics where software may require
  technical expertise to first compile and install before being used. Plugins
  can also be locked to specific versions to prevent backwards incompatible
  software changes breaking an existing workflow.}

@@

Genomer is written for the command-line to use in shells such as bash or zsh.
Command line tools are simple to automate using scripting, allowing
reproducible genome finishing and sharing genome projects between researchers.
For instance, during our microbial genome project we automated Genomer using
GNU Make build files, allowing the finishing steps to be repeated automatically
using the \verb+make+ command. This is an advantage of command-line tools
versus GUI-based tools, the latter are harder to automate and share steps.
Therefore Genomer may be of particular interest to bioinformaticians who prefer
automated approaches and scripting on the command line.

Additionally as Genomer uses plain text it is easy to store versions of the
project using a revision control system such as git \cite{git-scm}. This allows
tracking changes to the project and reverting errors back to earlier versions.
Example build files from our genome projects can be found on GitHub for a
simple plasmid \cite{plasmid-github} and a more complex genome
\cite{genome-github} sequence.

We have released Genomer as open source on GitHub \cite{genomer-github} and
have prepared documentation and video tutorials at http://next.gs. As we have
used this tool extensively to simplify our own genome projects we believe
genomer may also be useful to others in the field. We will illustrate the
possible application of genomer with an example use case.

@@

  \item{A genome scaffold is written in the scaffolder file format
  \cite{barton2012}. This specifies the order and orientation of contigs in the
  scaffold. Any unresolved gap regions can also be specified in this file.
  Paired-end sequencing or comparison to reference genomes are a source of
  contig orientation and order. In our genome projects we determine contig
  order by aligning contigs to a reference genome using nucmer
  \cite{kurtz2004}. A detailed overview of the scaffolder file format can be
  found on the genomer website.}

@@

  \item{When researchers are satisfied with the status of their draft the
  required files for submission to GenBank can be generated: assembled FASTA,
  contig FASTA, annotation tables, and scaffold layout in AGP. Additional
  options are used to add the required prefix to gene annotation IDs and begin
  locus ID numbering at the origin of the sequence. Genomer does not produce
  the .asn file which must be submitted to GenBank, but instead the files that
  may be used as the input to tbl2asn or sequin. }

@@

Genomer simplifies the process of generating the annotation table file required
for submission to GenBank where only gene entries are required. Specifically a
user need only provide a GFF3 file of annotations and Genomer can automatically
generate the additional encoded protein features. This however assumes no
intron/exon structure and therefore users wishing to submit annotations with
alternative splicing cannot take advantage of this genomer feature.

>>>>

Manually editing a genome sequence or annotation is a non-trivial problem
requiring effort on a researcher's part. The large size of FASTA sequence files
or general feature format (GFF) annotation files makes changing these in a text
editor difficult. Tasks such as organising project files, adding additional
sequences to an assembly, updating incorrect annotations or preparing files for
database deposition all divert effort from the main goal of analysing a genome.
As a result there is a steep learning curve for researchers embarking on genome
sequencing for the first time, while still requiring effort for experienced
researchers.

@@

During the process of sequencing several \emph{Pseudomonas fluorescens} strains
we developed a tool, called ``Genomer,'' to ease and simplify many of the
common steps in our genome project. While several graphical user interface
tools are available to work with genome projects \cite{tanaka2006,
wilkinson2002, lopez2011, carver2012, gordon1998}, Genomer instead provides a
simple command line interface. This allows scriptable automation and
reproducible results. We have open-sourced this tool under the MIT license and
here we describe its application.

@@

Genomer uses these two libraries to create a plugin system where third-party
software can be written and included. Plugin creators need only to prefix the
name of their plugin with \verb+genomer-plugin-+ and upload this to
rubygems.org or make the plugin available as a public git repository. A plugin
is specified within a Genomer project in a file named `Gemfile.' The
\verb+bundle update+ command will then automatically download and install the
required plugins plus any gem dependencies these gems rely on.

Genomer is implemented as a command line tool and tested on both Mac OS X and
Linux systems. The latest version of Genomer can installed from the Rubygems
package management system with the command \verb+gem install genomer+. Genomer
is based on our previously described ``Scaffolder'' software \cite{barton2012}
and is invoked on the command line with the command \verb+genomer+. Genomer
provides extensive documentation via UNIX manual pages and on the Genomer
website at http://next.gs.

@@

  \item[Mapping of annotations onto the assembled sequence.]{Genomer maps the
  coordinates of contig annotations to their respective positions in the draft
  genome sequence. This allows the scaffold file to be continuously improved
  and updated even after the contigs have been annotated. For instance,
  additional PCR sequences can be used to close gaps in the assembly and the
  location of all annotations be correspondingly updated.}

  \item[Generation of files for submission to GenBank.]{Submitting a genome
  sequence to GenBank database requires generating specific files. These may
  include a FASTA file of the draft sequence, a table of annotations, a FASTA
  file containing the individual contig sequences, and `a golden path' (AGP)
  \cite{agp-spec} file describing the placement of contigs. Genomer automates
  the generation of all these files from the scaffold file and corresponding
  GFF3 file and removes the effort to produce these manually.}

  \item[A stable interface and streamlined install process]{Genomer provides a
  plugin system built using the existing Ruby package management system:
  RubyGems and Bundler. This allows Genomer and its plugins to be automatically
  installed without requiring manual downloading and compilation by the user.
  This eliminates a common problem in bioinformatics where software may require
  technical expertise to first compile and install before being used. Plugins
  can also be locked to specific versions to prevent backwards incompatible
  software changes breaking an existing workflow.}

@@

Genomer is written for the command-line to use in shells such as bash or zsh.
Command line tools are simple to automate using scripting, allowing
reproducible genome finishing and sharing genome projects between researchers.
For instance, during our microbial genome project we automated Genomer using
GNU Make build files, allowing the finishing steps to be repeated automatically
using the \verb+make+ command. This is an advantage of command-line tools
versus GUI-based tools, is that the former are simpler to automate and share
steps with other researchers. Therefore Genomer may be of particular interest
to bioinformaticians who prefer automated approaches and scripting on the
command line.

Additionally as Genomer uses plain text it is easy to store versions of the
project using a revision control system such as git \cite{git-scm}. This allows
tracking changes to the project and reverting errors back to earlier versions.
Example versioned build files from our genome projects can be found on GitHub
for a simple plasmid \cite{plasmid-github} and a more complex genome
\cite{genome-github} sequence.

We have released Genomer as open source on GitHub \cite{genomer-github} and
have prepared documentation and video tutorials at http://next.gs. As we have
used this tool extensively to simplify our own genome projects we believe
Genomer may also be useful to others in the field. We will illustrate the
possible application of Genomer with an example use case.

@@

  \item{A genome scaffold is written in the scaffolder file format
  \cite{barton2012}. This specifies the order and orientation of contigs in the
  scaffold. Any unresolved gap regions can also be specified in this file.
  Paired-end sequencing or comparison to reference genomes are a source of
  contig orientation and order. In our genome projects we determine contig
  order by aligning contigs to a reference genome using nucmer
  \cite{kurtz2004}. A detailed overview of the scaffolder file format can be
  found on the Genomer website.}

@@

  \item{When researchers are satisfied with the status of their draft the
  required files for submission to GenBank can be generated: assembled FASTA,
  contig FASTA, annotation tables, and scaffold layout in AGP. Additional
  options are used to add the required prefix to gene annotation IDs and begin
  locus ID numbering at the origin of the sequence. Genomer does not produce
  the .asn file which is submitted to GenBank, but instead the files that may
  be used as the input to tbl2asn or sequin. }

@@

Genomer simplifies the process of generating the annotation table file required
for submission to GenBank where only gene entries are required. Specifically a
user need only provide a GFF3 file of annotations and Genomer will generate the
additional encoded protein features. This however assumes no intron/exon
structure and therefore users wishing to submit annotations with alternative
splicing cannot take advantage of this Genomer feature.
